---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Yike Ma"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

**!! AI generated code should NOT be used in this assignment**


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
getwd()
library(tidyverse)
library(agricolae)
library(lubridate)
raw <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = T)
raw$sampledate <- mdy(raw$sampledate)
#2
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer: 
H0: There are no relationship between lake temperature and depth, and mean lake temperature is equal across all lakes.
Ha: There is a linear relationship such that lake temperature decreases with lake depth.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
ntl <- raw %>% 
  filter(month(sampledate) == 7) %>% 
  select(c(lakename:temperature_C), -sampledate) %>% 
  na.omit()

summary(ntl)
#5
temp_dep <- ntl %>% 
  ggplot(aes(x = depth, y = temperature_C))+
  geom_point()+
  geom_smooth(method = "lm")+
  ylim(c(0,35))+
  xlab("Depth (m)")+
  ylab(expression("Temperature"*~degree*"C"))
temp_dep
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The temperature of the lake decreases with depth. The linear regression model suggests there is a negative linear trend, however, there is a deviation of data points when the depth is above 10m.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
lm <- lm(data = ntl, temperature_C~depth) 
summary(lm)
```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:
The variation of the linear model (R-squared) is 0.7387. F value is 2.75e+04 and p-value < 2.2e-16, indicating this model is statistically significant.
The intercept of the linear regression model appears to be 21.96 and a slope of -1.95, indicating every 1m deeper in depth results in 1.95 degree Celcius of decrease.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
mlr <- lm(data = ntl, temperature_C~year4+daynum+depth)
step(mlr)

#10
mlr <- lm(data = ntl, temperature_C~year4+daynum+depth)
summary(mlr)
```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: 
AIC suggests to use year4, daynum and depth as variables to predict the multiple linear regression model.
The variance is 0.7412 here, which demonstrates and improvement than single linear regression model(0.7387).


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
temp_anova <- aov(data = ntl, temperature_C ~ lakename)
temp_anova
summary(temp_anova)

temp_lm <- lm(data = ntl, temperature_C ~ lakename)
temp_lm
summary(temp_lm)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: We obtained a p value of <2e-16 here, which rejects the null hypothesis, meaning mean temperature are significanly different among the lakes. But we cannot tell which lake is significantly different to another merely running anova test.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
ntl_lake <- ntl %>% 
  ggplot(aes(x = depth, y = temperature_C, colour = lakename))+
  geom_point(alpha = 0.5, size = 1)+
  geom_smooth(method = "lm", se = F)+
  ylim(c(0,35))+
  xlim(c(0,15))+
  xlab("Depth (m)")+
  ylab(expression("Temperature"*~degree*"C"))+
  labs(color = "Lake Name")
ntl_lake
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(temp_anova)

ntl.groups <- HSD.test(temp_anova, "lakename", group = TRUE)
ntl.groups
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:
Paul Lake and Ward Lake statistically share same mean temperature as Peter lake.
Mean temperature of Central Long Lake and Crampton Lake are significantly different from others. However, there is no lake statistically distinct from all the other lakes in this example.

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We can run two sample t-test.



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
ntl.Crampton.Ward <- ntl %>% 
  filter(lakename %in% c("Crampton Lake" , "Ward Lake"))

t.test(ntl.Crampton.Ward$temperature_C~ntl.Crampton.Ward$lakename)
```

>Answer: T-test suggests a p-value of 0.26 (<0.5), so the mean temperature in both lakes are different. 
Mean temperature for Crampton Lake and Ward Lake is 15.35 and 14.46 degree Celcius, respectively. However, the mean values were shown statistically the same in question 16, which does not match, this could be caused by the larger categorical variables in question 16.

